---
title: "Visualizing fish movements with animated videos in R: The ggplot2 and gganimate packages"
subtitle: "Workshop Handout - GLATOS, February 28, 2019"
author: "Todd Hayden (haydento@msu.edu), Mike Lowe"
date: "Updated: `r Sys.Date()`"
output:
  rmarkdown::html_document:
    theme: default
  rmarkdown::pdf_document:
    toc: false
---

# Introduction

The `glatos` package includes several functions that facilitate the
creation of animated videos of animal movements from acoustic
telemetry data. The primary functions for building animated videos are
the `interpolate_path` and `make_frames` functions.
`interpolate_path` calculates interpolated positions of animals
between detections and `make_frames` creates a series of snapshots of
interpolated and detected fish locations over time from the output of
`interpolate_path`. In addition to creating and outputting frames,
`make_frames` can also stitch frames together to create an animated
video of movements using the open source FFmpeg software.  Although
`make_frames` was developed to allow customization of the output
content and layout of frames, some may find the level of customization
to be insufficient for their needs. 

In this vignette, we present a short example of an alternative approach for creating animations using the `ggplot2` and `gganimate` R packages. ^[For more information about
obtaining `gganmiate` and `ggplot2`, see https://gganimate.com/ and
https://ggplot2.tidyverse.org/] `ggplot2` is a popular and powerful
package for plotting figures based on The Grammer of Graphics approach
to visualizing data and developed for R by Hadley Wickham.  ^[Leland
Wilkinson,https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448]
`ggplot2` differs from base plotting functions in R by its layered
approach where graphical output is defined by combing layers of plot
objects.  

In `ggplot2` variables in data are mapped to aesthetic
categories to build the plot.  In contrast, calls to base R plotting
functions are rendered immediately after they are submitted. The
`ggplot2` framework enables complex plots to be specified with few
lines of code however in our experience, customized `ggplot2` figures
often require substantial effort.  Base R plotting functions often
require extra effort to create but are easier customize.  Both
frameworks will produce publication-quality graphical output and have
large user-bases. 

The `gganimate` package is an extension to
`ggplot2` that provides support for animated graphical output ^[see
https://gganimate.com/articles/gganimate.html].  In the following
code, we use the `make_transition` function on an example dataset
included in `glatos` to create a custom transition layer necessary for
interpolating fish movements.  We then use the `interpolate_path`
function in `glatos` to interpolate fish positions in space and time
^[See `glatos` package help for more information about functional
arguments and outputs of `make_frames`, `make_transition`, and `interpolate_path`. Additional details about creating animated videos
of fish movements using the `glatos` package is found in the workshop
manual "A guide to R for analyzing Acoustic Telemetry Data V2.0".  A
pdf of manual can be downloaded here
https://github.com/ocean-tracking-network/glatos/wiki/Past-R-workshops-and-manuals].Finally,
we end the vignette with a demonstration of code to create an
animation of fish movements using `ggplot2` and `gganimate`.

In the first chunk we load necessary packages, obtain example data
from `glatos`, clean up the data, and create a figure of lakes Huron
and Erie (Figure 1). The `data.table` R package is used for data
manipulations in this vignette to improve efficiency. ^[Data.table
information: https://github.com/Rdatatable/data.table/wiki]

```{r, echo=FALSE}
fig.cap1 <- "Lakes Huron and Erie extracted from Great Lakes polygon included in the `glatos` package"

```
```{r, echo=TRUE,fig.margin=TRUE,fig.cap=fig.cap1, fig.height=8,fig.width=8,tidy=TRUE,warning=FALSE,message=FALSE}

library(glatos)
library(sp)
library(sf)
library(raster)
library(gganimate)
library(ggplot2)
library(data.table)

# get polygon of the Great Lakes 
data(greatLakesPoly)

# convert to sf spatial object
lakes <- st_as_sf(greatLakesPoly)

# crop polygon to extract Lake Huron and Lake Erie
lakes <-  st_crop(lakes, xmin = -84.8, xmax = -79.6, ymin = 41.3, ymax= 46.6)
plot(st_geometry(lakes), col = "grey")

# get example walleye detection data
det_file <- system.file("extdata", "walleye_detections.csv",
                        package = "glatos")
det <- read_glatos_detections(det_file)

# bring in receiver file and convert to sf spatial object
rec_file <- system.file("extdata", "sample_receivers.csv", package = "glatos")
recs <- read_glatos_receivers(rec_file)
setDT(recs) # convert to data.table object- necessary for sf package
recs <- recs[deploy_long < -80,]
recs_sf <- st_as_sf(recs, coords = c("deploy_long", "deploy_lat"), crs = 4326) 

```

In the next bit of code, we create a transition layer from the `lakes`
object. The transition layer is needed to calculate non-linear interpolated
movement paths that avoid impossible overland movements of fish. We
set the `all_touched` argument to true (`all_touched = TRUE`) to allow any 
pixel that touches the polygon outline of the Great Lakes to be coded as water in the
resulting transition layer.  If `all_touched` is false (`all_touched = FALSE`), 
then more than half of the pixel must be in the polygon in order to be coded as
water.


```{r, cache=FALSE,tidy=TRUE} 

# temporarily convert sf spatial to sp
# make_transition only accepts sp SpatialPolygonsDataFrame
lakes <- as(lakes, "Spatial")

# make transition layer.
tran <- make_transition(lakes, res = c(0.01, 0.01), all_touched = TRUE)

# convert back to sf
lakes <- st_as_sf(lakes)

```

```{r, echo=FALSE,tidy=TRUE}

cap2 <- "Plot of transition layer, receivers (orange circles), fish detection location (red circles) in lakes Huron and Erie." 

```

```{r, fig.width=8,fig.height=8,fig.margin=TRUE,fig.cap=cap2,tidy=TRUE}


# create quick plot of fish and receivers to make sure they are "on the map" (Figure 2)
unique_fish <- unique(det, by = c("deploy_lat", "deploy_long")) 
plot(tran$rast)
plot(st_geometry(lakes), add = TRUE)
plot(st_geometry(recs_sf), pch=20, col= "orange", add = TRUE)
points(unique_fish$deploy_long, unique_fish$deploy_lat, pch = 20, col = "red")

```

Next we interpolate detection data and do some cleanup.


```{r, eval = FALSE,tidy=TRUE}

# interpolate detections.
pos2 <- interpolate_path(det, trans = tran$transition, lnl_thresh=0.99)

```

```{r, echo = FALSE,tidy=TRUE,message=FALSE}

dump <- capture.output( {pos2 <-interpolate_path(det, trans = tran$transition, lnl_thresh=0.99)})

```

```{r, echo=TRUE,tidy=TRUE}

# convert pos2 to data.table
setDT(pos2)

# extract unique bin_timestamp from the interpolated data
int <- unique(pos2, by = "bin_timestamp")

```

In the next code chunk, we use a `data.table` join to add the unique
timestamps extracted from `pos2` in the last line of code above (named
`int`) to the receiver object.  We use receiver deployment and
recovery timestamps to identify if the receiver was active in that
`bin_timestamp`.  In other words, we need to know what receivers were
deployed during each `bin_timestamp` interval so we can display
receiver deployment and recovery in the animation.  This step is
really easy with `data.table` and would be a bit complicated using
base R functions.

```{r, echo=TRUE,tidy=TRUE}

# Add bin_timestamp to receivers, based on deploy/recover timestamps.
# Removes unnecessary columns in output to simplify
# This is a data.table non-equi join...
recs <- recs[int, .(deploy_lat = x.deploy_lat,
                    deploy_long = x.deploy_long,
                    station = x.station,
                    deploy_date_time = x.deploy_date_time,
                    recover_date_time = x.recover_date_time,
                    bin_timestamp = i.bin_timestamp),
             on = .(deploy_date_time <= bin_timestamp,
                    recover_date_time >= bin_timestamp),
             allow.cartesian = TRUE]


```

We are ready to create the animation.  The first step is to create a
static image of all receivers, detections, and interpolated detections
with `ggplot2` (all frames combined).  We changed the color of fish
detections to red circles, interpolated positions to blue circles and
made detections larger than interpolated detections.

```{r, echo=FALSE,tidy=TRUE}

cap3 <- "Static figure of fish detections (red), interpolated positions (blue), and receivers (orange) plotted using `ggplot2`"
```

```{r,fig.width=8,fig.height=8,fig.cap=cap3,message=FALSE}

ggplot(data = lakes) +
  geom_sf(color = "black", fill = "lightgrey") +
  geom_point(data = recs, aes(x = deploy_long, y = deploy_lat), size = 2, color = "orange", inherit.aes = FALSE) +
  geom_point(data=pos2, aes(x=longitude, y=latitude, group=animal_id, color=record_type, size=record_type), inherit.aes=FALSE) +
  xlab("Longitude") +
  ylab("Latitude") +
  scale_color_manual(values=c("red", "blue")) +
  scale_size_manual(values=c(2,1))

```

Now we will animate the plot with `gganimate`.  We repeat the code
from the previous code chunk and add an additional layer to create the
animation and a layer to display a clock (`transition_time`, `ggtitle`).


```{r, echo = FALSE,tidy=TRUE,message=FALSE}

cap4 <- "Animated figure of fish movements in lakes Huron and Erie.  Only last image of animation is included in this vignette."

```


```{r,fig.width=8,fig.height=8,fig.cap=cap4,message=FALSE,warning=FALSE,fig.keep="last"}

ggplot(data = lakes) +
  geom_sf(color = "black", fill = "lightgrey") +
  geom_point(data = recs, aes(x = deploy_long, y = deploy_lat),
             size = 2, color = "orange", inherit.aes = FALSE) +
  geom_point(data=pos2, aes(x=longitude, y=latitude,
                            group=animal_id, color=record_type,
                            size=record_type), inherit.aes=FALSE) +
  xlab("Longitude") +
  ylab("Latitude") +
  scale_color_manual(values=c("red", "blue"))+
  scale_size_manual(values=c(2,1)) +
  transition_time(bin_timestamp) +
  ggtitle('{frame_time}')

```

As mentioned previously, the addition of a `transition_time` and
`ggtitle` component in the code above creates the animation.  The
resulting animation may be customized (number of frames, duration of
animation, frames per second, etc) by specifying arguments to the
`animate` function.  Please see `?animate` for more information.
Also, specifying `anim_save` may be used to write the animation object
to disk.

This vignette provides a basic, high-level summary of creating an
animated video with `ggplot2` and `gganimate`.  These packages
provide much more functionality than included in this vignette.
Please refer to documentation for `gganimate` and `ggplot2` for additional
examples and customization that may be useful for visualizing fish
movements.



